\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-04 coqimpl.tex
Fail to move some parts from sec 3.2 to sec 3.1.

Inspired by the above challenges and Family Polymorphism, we propose
the following extension to Coq's Vernacular Commands, and the corresponding STLC example.
% basically write down the specification of surface Coq commandline
%   and then explain  how the language design makes the programmer’s life easier
\begin{align*}
  V := \quad &"Family"~\langle ident\rangle~["extends"~\langle ident\rangle] 
  \quad |\quad "EndFamily" \\
  & |\quad "FRecursor"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle~"using"~\langle ident\rangle \\
  & | \quad "FTheorem"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle \\
  & |\quad "Extend"~"Family"~\langle ident\rangle \\ 
  & |\quad ["Extend"]~"FInductive"~\langle ident\rangle \\ 
  & |\quad ["Final" | "Override"]~"Field"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\
  & | \quad "FieldClaim"~\langle ident\rangle~":"~\langle term \rangle \\
  & | \quad "Inherit"~\langle ident\rangle \quad | \quad "Inherit Until"~\langle ident\rangle \quad | \quad ...
\end{align*}

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
 FInductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
 FInductive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term ...
 Family subst_internal.
  Final Field var := ..
  Final Field lam := ..
   ...
 EndFamily.
 FRecursor subst
  about tm motive ..
  using subst_internal.
 ...
 FInductive has_type 
  : context -> term -> type := ...
 FTheorem subst_lemma 
  about has_type 
  motive ...
  ..
 Qed.
 FInductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC_bool extends STLC.
 Extend FInductive ty : Set :=
  | bool : ty.

 Extend FInductive term : Set := 
  | tt : term | ff : term 
  | tif : term -> term -> term -> term
 Extend Family subst_internal.
  Final Field tt := ..
  Final Field ff := ..
  Final Field tif := ..
 EndFamily. (* Automatic inheriting others 
              from STLC.subst_internal *)
 Inherit subst.
(* Plugin will redo exhaustiveness check, in case 
 subst_internal is not extended correctly *)
 Extend FInductive has_type (* .. and new ty too *)
  : context -> term -> type := ...
 Extend FTheorem subst_lemma.
  ... (* Need to prove extra cases *)
 Qed.

@@

These are the core extensions we proposed to support \cref{fig:STLC-example}. We use "Family" and "EndFamily" to scope, and "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.} Both "Field" and "FieldClaim" are only allowed to refer to the former fields, to avoid possible inconsistency mentioned in \ref{chg:consistency}.
Nested families are also allowed.

@@

We can inherit a top-level family "parent" using \mintinline{Coq}{Family
... extends parent} as a new top-level family. As expected, "Override"
will override a field from the parent. "Inherit" can be used to inherit
nested families, inductive types, and fields. The fields inside a family
are order-sensitive because we have dependent types here. So "Inherit
Until" is used if the user wants to inherit a range of fields.  "Extend Family" is used only inside a inheriting family. Thus nested
family can only be inherited or extended together with its belonging family
(i.e. ultimately top level family), thus our support of family polymorphism is
incomplete.

% This incomplete support can already be useful for extending functions.
% Take \cref{fig:plugin-example2} as example, if we want to extend boolean
% inductive type and its negation, we only need to specify the new
% constructor for boolean and how negation deals with the new constructor.
% As a result, the programmer does not have to duplicate code---otherwise
% the user will need to copy and paste the whole original negation
% definition and add a new pattern matching case.

@@

\textbf{Reasoning Needs Non-Overridable.}
This compilation strategy
has already allowed fields to be type-dependent on former fields thus we can see concretely how
\ref{chg:definition-relevant-reasoning} appears in our current formulation and what we needs to solve this challenge.

@@

To resolve this, we provide \textit{family sealing} mechanism to simulate the sigma type. In a sealed family, every field will expose its definition inside the family but a family as a whole will be sealed into a ``relaxed'' type. Thus we can override one sealed family as a whole and change the internal implementation.

@@

However, we later still need the concrete eliminator "b" for doing exhaustiveness checking (as in \ref{chg:extensible-inductive-type}), thus we will store the corresponding info somewhere.  Here we show one exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module "b_rec_12". This concrete eliminator will be used by our "FRecursor". For example, in \cref{fig:plugin-example2}, "FRecursor" command will feed Coq a module "v_33_34" for doing type-checking with specified handlers (e.g. "neg_handler") during compilation---what it does is basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq. Note that, at this point of the example, "FRecursor" will only do type check. The compilation of "FRecursor" into a component of the compiled module only happens when closing the whole family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
  Definition neg := b_rec (fun _ => b) neg_handler.tt neg_handler.ff.
\end{minted}
where "b_rec" and "b" are the corresponding recursor and the (vanilla) inductive type in that compiled module. Thus a "FRecursor" field is unlike other fields, even though it is exposed as a mundane function for the following fields.

Importantly, the recursor field (e.g. "neg") is not actually inheritable, because it is not compiled like any other fields---recall how other fields are compiled into reusable pieces of (parametrized) modules, a recursor field is not compiled like that. For all the children family, we have to reconstruct and override the "FRecursor" field with a ``new'' recursor using (maybe the same) handler family. This is affordable since each aggregated recursor handler doesn't need
to be rechecked---look at "B2.neg_handler", there is only one new
handler for "uu" and the handlers (mundane fields) from the parent
"B.neg_handler.tt,B.neg_handler.ff" have already been type-checked and
inherited (since they only rely on ``the abstracted
interface'' module "b_3"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "ff_24"---doesn't require another type-check and is re-used during the type-checking and compilation of "B2.neg". For example, for the new "neg", when compiling family "B2" into a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
  Definition neg := b_rec (fun _ => b) neg_handler.tt neg_handler.ff neg_handler.uu.
\end{minted}
where "b_rec" and "b" are the corresponding recursor and the (vanilla but extended and modified) inductive type in that compiled module. Note that, even though "b_rec" and "b" has the same name as before, they are in the newly-compiled module so they refer to totally different things.

>>>>

Inspired by the above challenges and Family Polymorphism, we propose
the following extension to Coq's Vernacular Commands, and the corresponding STLC example.
% basically write down the specification of surface Coq commandline
%   and then explain  how the language design makes the programmer’s life easier
\begin{align*}
  V := \quad &"Family"~\langle ident\rangle~["extends"~\langle ident\rangle] 
  \quad |\quad "EndFamily" \quad | \quad "SealFamily" \\
  & |\quad "FRecursor"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle~"using"~\langle ident\rangle \\
  & | \quad "FTheorem"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle \\
  & |\quad ["Extend" | "Override"]~"Family"~\langle ident\rangle \\ 
  & |\quad ["Extend"]~"FInductive"~\langle ident\rangle \\ 
  & |\quad ["Final" | "Override"]~"Field"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\
  & | \quad "FieldClaim"~\langle ident\rangle~":"~\langle term \rangle \\
  & | \quad "Inherit"~\langle ident\rangle \quad | \quad "Inherit Until"~\langle ident\rangle \quad \\
  & | \quad "Closing Fact"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\ 
  & | \quad "Closing Fact"~\langle ident\rangle~":"~\langle term \rangle~"by"~"\{"\langle \text{one tactic expression}\rangle"\}" \\ 
  & | \quad "MetaData"~\langle ident\rangle \quad | \quad "EndMetaData" \\ 
\end{align*}

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
 FInductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
 FInductive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term ...
 Family subst_internal.
  Final Field var := ..
  Final Field lam := ..
   ...
 EndFamily.
 FRecursor subst
  about tm motive ..
  using subst_internal.
 ...
 FInductive has_type 
  : context -> term -> type := ...
 FTheorem subst_lemma 
  about has_type 
  motive ...
 Proof.  ...
 Qed.
 FInductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC_bool extends STLC.
 Extend FInductive ty : Set :=
  | bool : ty.

 Extend FInductive term : Set := 
  | tt : term | ff : term 
  | tif : term -> term -> term -> term
 Extend Family subst_internal.
  Final Field tt := ..
  Final Field ff := ..
  Final Field tif := ..
 EndFamily. (* Automatic inheriting others 
              from STLC.subst_internal *)
 Inherit subst.
(* Plugin will redo exhaustiveness check, in case 
 subst_internal is not extended correctly *)
 Extend FInductive has_type (* .. and new ty too *)
  : context -> term -> type := ...
 Extend FTheorem subst_lemma. Proof.
  ... (* Need to prove extra cases *)
 Qed.

@@

These are the core extensions we proposed to support \cref{fig:STLC-example}, and we adapt them into \cref{fig:STLC-example2}. We use "Family" and "EndFamily" to scope. "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.} Both "Field" and "FieldClaim" are only allowed to refer to the former fields, to avoid possible inconsistency mentioned in \ref{chg:consistency}.
Nested families are also allowed.

@@

We can also use "Family" and "SealFamily" to scope a (nested) family. The resulting family only support direct inheritance or "Override" (i.e. we can only override a sealed family as a whole together)---we will later explain this design decision and its usage, especially on resolving \ref{chg:definitional-relevant-reasoning}. What's more, "Override" family can only apply to sealed family as well (not allowed on normal family).

We can inherit a top-level family "parent" using \mintinline{Coq}{Family ... extends parent} as a new top-level family. As expected, "Override"
will override a field from the parent. "Inherit" can be used to inherit
nested families, inductive types, and fields. "Final" is used to claim that the decorated field is non-overridable, and later we will show how it is used to resolve\ref{chg:definitional-relevant-reasoning}. The fields inside a family
are order-sensitive because we have dependent types here. So "Inherit
Until" is used if the user wants to inherit a range of fields.  "Extend Family" is used only inside a inheriting family. Thus nested
family can only be inherited or extended together with its belonging family
(i.e. ultimately top level family), thus our support of family polymorphism is
incomplete.

We will explain the usage of "Closing Fact" and "MetaData" later---they are here to resolve \ref{chg:software-engineering} to some extent.
% This incomplete support can already be useful for extending functions.
% Take \cref{fig:plugin-example2} as example, if we want to extend boolean
% inductive type and its negation, we only need to specify the new
% constructor for boolean and how negation deals with the new constructor.
% As a result, the programmer does not have to duplicate code---otherwise
% the user will need to copy and paste the whole original negation
% definition and add a new pattern matching case.

@@

\textbf{Reasoning Needs Non-Overridable.}
This compilation strategy
has already allowed fields to be type-dependent on former fields thus we can see concretely how
\ref{chg:definition-relevant-reasoning} appears in our current formulation and how "Final" can help to solve this challenge.

@@

To resolve this, we provide \textit{family sealing} mechanism to simulate the sigma type. In a sealed family, every field will expose its definition inside the sealed family but a family as a whole will be sealed into a ``relaxed family type''. Thus we can override one sealed family as a whole and change the internal implementation as long as they can satisfy the ``relaxed family type''.

@@

However, we later still need the concrete eliminator "b" for doing exhaustiveness checking (as in \ref{chg:extensible-inductive-type}), thus we will store the corresponding info somewhere.  Here we show one exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module "b_rec_12". This concrete eliminator will be used by our "FRecursor". For example, in \cref{fig:plugin-example2}, "FRecursor" command will feed Coq a module "v_33_34" for doing type-checking with specified handlers (e.g. "neg_handler") during compilation---what it does is basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq. Note that, at this point of the example, "FRecursor" will only do type check. The compilation of "FRecursor" into a component of the compiled module only happens when closing the whole family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition neg := b_rec (λ _ ⇒ b) neg_handler.tt neg_handler.ff.
\end{minted}
where "b_rec" and "b" are the corresponding recursor and the (vanilla) inductive type in that compiled module. Thus a "FRecursor" field is unlike other fields, even though it is exposed as a mundane function for the following fields.

Importantly, the recursor field (e.g. "neg") is not actually inheritable, because it is not compiled like any other fields---recall how other fields are compiled into reusable pieces of (parametrized) modules, a recursor field is not compiled like that. For all the children family, we have to reconstruct and override the "FRecursor" field with a ``new'' recursor using (maybe the same) handler family. This is affordable since each aggregated recursor handler doesn't need
to be rechecked---look at "B2.neg_handler", there is only one new
handler for "uu" and the handlers (mundane fields) from the parent
"B.neg_handler.tt,B.neg_handler.ff" have already been type-checked and
inherited (since they only rely on ``the abstracted
interface'' module "b_3"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "ff_24"---doesn't require another type-check and is re-used during the type-checking and compilation of "B2.neg". For example, for the new "neg", when compiling family "B2" into a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
Definition neg := b_rec (λ _ ⇒ b) neg_handler.tt neg_handler.ff neg_handler.uu.
\end{minted}
where "b_rec" and "b" are the corresponding recursor and the (vanilla but extended and modified) inductive type in that compiled module. Note that, even though "b_rec" and "b" has the same name as before, they are in the newly-compiled module so they refer to totally different things.
